---
title: "MA684 Discussion"
author: "Yitong"
date: "September 06, 2016"
output: html_document
---

<style type="text/css">
body{
  font-family: Helvetica;
  font-size: 12pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set
library(ggplot2)
library(knitr)
```

<p></br></p>

# Introduction

---

For homework 1 you will create your first knitr report and perform bunch of exploratory data analysis. This will be the first of many reports you will generate throughtout the year.  Many questions will be open ended but you will be evaluated on the

1. Accuracy and creativity of your answer.
2. Clenlyness of your report.
3. How legible your code is written.

There are couple rules that you must follow

1. Put comments on your code where ever necessary.
2. The code should run from top to bottom.
3. Name the variables well so that what you are trying to do is obvious.
4. Test your script when you are done.

Make sure the code you submit runs on a clean R environment.
If a grader runs your code and get an error, you get 50% off for that homework after fixing the mistake.

You are welcome to help each other but write the code on your own 
since we will test you on your coding ability here and there.
These spot tests are to get you prepared for your future interviews.

**About this document** </p> 
In this discussion, you will learn how to use R Markdown with `knitr` and  create a nice report.</p>

The base structure for a markdown file contains both R code and Knitr code. Any code between the sets of three backticks is R code and any code outside is Knitr.</p>

Chunk options are written in one line. The followings are some popular chunk options. </p>
- `echo=TRUE` - display the command  ;  `error=FALSE` - do not display any error messages 
- `eval=TRUE` - run the code         ;  `message =FALSE` - do not display any messages 
- `include=TRUE` - show the output   ;  `warning =FALSE` - do not display any warning  messages 

<p></br></p>

# Exploratory Data Analysis

---

You can get the example data from this [link](http://socserv.socsci.mcmaster.ca/jfox/Books/Applied-Regression-3E/datasets/index.html) </p>

The code to read in the data is already written for you. Your job is to look at the data and think about the following

1. What are some hypothesis that can be addressed from the data?
2. Does regression seem like a good analysis strategy?
3. Is there some transformation that you can apply so that you can use regression?
<p></br></p>

## <span style="color:blue"> Example: United Nations Social Indicators Data </span>

---

### Variables: 
  - region: Africa, America, Asia, Europe, Oceania
  - educationMale: Average number of years of education for men
  - lifeMale: Expectation of life at birth for men
  - lifeFemale: Expectation of life at birth for women
  - infantMortality: infant deaths per 1000 live births.

### Making plots

For a quick plot, you can use qplot() which wraps up the details of ggplot. The syntax looks like this </p>
`qplot(x, y, data=, color=, shape=, size=, geom=, method=, ...)` 

The `ggplot2` package provides a more flexible syntax for creating elegant and complex plots. </p>
`ggplot(data, mapping) + layer( stat = "", geom = "", position = "", geom_parms = list(),...)` 

Now let's use the example data to do some data visualization.

<p></br></p>
**Scatterplot**

```{r, message=FALSE, warning=FALSE}
# read data
unitednations<-read.table("http://socserv.socsci.mcmaster.ca/jfox/Books/Applied-Regression-3E/datasets/UnitedNations.txt")

# scatterplot using qplot()
qplot(lifeFemale, lifeMale, data = unitednations, colour = region, main = "scatterplot using qplot()" )
```

Add a smoother to show a trend

```{r,  message=FALSE, warning=FALSE}
qplot(lifeFemale, lifeMale, data = unitednations, colour = region, geom = c("point", "smooth"))   
```

Add a line specified by slope and intercept by using `geom_ablin` in ggplot()

```{r, message=FALSE, warning=FALSE, fig.width=7, fig.height=5, fig.align='left'}
# scatterplot using ggplot()
unitednations_gp <- ggplot(unitednations)

unitednations_gp + aes(x=lifeFemale, y=lifeMale, color=region)+ geom_point()+ 
                   geom_abline(intercept=0,slope=1) + xlab("Expectation of life at birth for women") + 
                   ylab("Expectation of life at birth for men") +
                   ggtitle("scatterplot using ggplot()")
```

Split up your data by one or more variables - lay out panels in a grid using `facet_grid` </p>
e.g. facet panels by region (a categorical variable with 5 levels) and show regions on the top
```{r, message=FALSE, warning=FALSE, fig.width=13, fig.height=7}
unitednations_gp +aes(x=lifeFemale, y=lifeMale, color=region)+geom_point()+ 
                  stat_smooth(method=lm)+
                  facet_grid( .~region, scales="fixed", labeller= label_both)
```

<p></br></p>

**Histogram and density plot**

```{r, message=FALSE, warning=FALSE}
# histogram
unitednations_gp + aes(x=lifeFemale, fill=region)+ geom_histogram(binwidth =1)
```

To compare distributions of "lifeFemale" by regions, we can use a density plot

```{r, message=FALSE, warning=FALSE}
# density plot
unitednations_gp + aes(x=lifeFemale,fill = region, colour = region)+ geom_density(alpha = .1)
```

If you want to display density plots by region, `facet_wrap` can wrap with a certain number of columns or rows. 
```{r,message=FALSE, warning=FALSE, fig.width=13, fig.height=8}
unitednations_gp + aes(x=lifeMale,fill = region, colour = region)+ geom_density(alpha = .1)+ 
                   facet_wrap(~region, nrow = 2)
```

<p></br></p>

### Fitting a model
For example, we fit a linear regression model with two predictors ("educationMale","region") and a response variable "lifeMale".
```{r}
regout = lm(lifeMale ~ factor(region) + educationMale, data =unitednations)
```

The kable function is sufficient to make a table for the  model output, although it doesn’t have many options.
```{r, kable}
kable(summary(regout)$coef, digits=2)
```

<p></br></p>

## <span style="color:blue"> Example: Leinhardt and Wasserman’s Data on Infant Mortality </span>

---

### Variables: 
 - income: Per-capita income in U.S. dollars. 
 - infant: Infant-mortality rate per 1000 live births.  
 - region: Africa; Americas; Asia, Asia and Oceania; Europe.  
 - oil: Oil-exporting country: no, yes.  
  
### Plots

Boxplot is good for measuring dispersion and telling outlier points. </p>
```{r, message=FALSE, warning=FALSE, fig.width=13, fig.height=7}
# read data
leinhardt<-read.table("http://socserv.socsci.mcmaster.ca/jfox/Books/Applied-Regression-3E/datasets/Leinhardt.txt")

# boxplot
ggplot(leinhardt) + aes(x=region, y=infant/1000, color=oil)+ geom_boxplot(width = .7)+
                    facet_wrap(~oil, ncol = 2)+ stat_summary(fun.y=mean, colour="darkred", geom="point", shape=15, size=2)+
                    ylab("infant rate") + ggtitle("Relationship between region and infant rate")
```

Seems like linear relationship only exists in Europe
```{r, message=FALSE, warning=FALSE, fig.width=13, fig.height=7}
# scatterplot
ggplot(leinhardt) + aes(x=income, y=infant, color=region) + geom_point()+ geom_smooth(method = "lm")+
                   facet_grid(. ~region)
```

Any luck if we take a log transformation for the data?
```{r, message=FALSE, warning=FALSE,fig.width=13, fig.height=7}
ggplot(leinhardt) + aes(x=log(income), y=log(infant), color=region)+ geom_point() +
                    geom_smooth(method = "lm")+ facet_grid(. ~region)
```

### Model
You can fit a linear model for data only in Europe
```{r}
logout = lm(infant ~ income, data = leinhardt[which(leinhardt$region=="Europe"),]) 
kable(summary(logout)$coef, digits=2)
```

How about the model for the whole data?
```{r}
logout2 = lm(log(infant) ~ log(income)*region, data = leinhardt)  
kable(summary(logout2)$coef, digits=c(2, 2, 3, 3))
```

<p></br></p>

## <span style="color:blue"> Example: Sahlins’s Data on Agricultural Production in Mazulu Village </span>

---

### Variables: 
  - consumers: weighted number of consumers per gardener in each household
  - acres: number of acres per gardener tended in each household

Notes: The observations are 20 households in an African village. </p>

### Plot
Draw a scatterplot of acres versus consumers. What kind of relationship can you see? linear or nonlinear?
```{r, message=FALSE, warning=FALSE}
# read data
sahlins<-read.table("http://socserv.socsci.mcmaster.ca/jfox/Books/Applied-Regression-3E/datasets/Sahlins.txt")

# use geom_smooth to help see patterns 
ggplot(sahlins) + aes(x=consumers, y=acres)+ geom_point()+ geom_smooth()
```

### Model
The relationship seems to "locally polynomial", so we use LOESS regression which stands for locally weighted scatterplot smoothing.
```{r}
regout2 = loess(acres ~ consumers, data =sahlins, span=.5)
summary(regout2)
```

<p></br></p>

Let's plot fitted values. A nonlinear line looks good in general.  
```{r}
plot(acres ~ consumers,data =sahlins)
i = order(sahlins$consumers)
lines(sahlins$consumers[i],regout2$fitted[i],col="red")
```

